---
title: "Multivariate extremes"
author: "Anna Kiriliouk and Johan Segers"
date: "UCLouvain and KU Leuven, October 2024"
output: pdf_document
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{bm}
  - \newcommand{\rbr}[1]{\left(#1\right)}
  - \newcommand{\cbr}[1]{\left\{#1\right\}}
  - \newcommand{\pr}{\operatorname{\mathsf{P}}}
  - \newcommand{\expec}{\operatorname{\mathsf{E}}}
  - \newcommand{\sig}{\sigma}
  - \newcommand{\evi}{\xi}
  - \newcommand{\bevi}{\bm{\evi}}
  - \newcommand{\stdf}{L}%{\ell}
  - \newcommand{\Pick}{A}
  - \newcommand{\exCoef}{\theta}
  - \newcommand{\expms}{\Lambda}
  - \newcommand{\angms}{\varPhi}
  - \newcommand{\angmspdf}{\phi}
  - \newcommand{\hstdf}{\widehat{\stdf}}
  - \newcommand{\hangms}{\widehat{\angms}}
  - \newcommand{\hbP}{\widehat{\bP}}
  - \newcommand{\hP}{\widehat{P}}
  - \newcommand{\hR}{\widehat{R}}
  - \newcommand{\hF}{\widehat{F}}
  - \newcommand{\hFj}{\widehat{F}_{j}}
  - \newcommand{\rank}{\operatorname{Rnk}}
  - \newcommand{\hU}{\widehat{U}}
  - \newcommand{\hpr}{\widehat{\pr}}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align="center")
```

```{r prelim}
library(mev)
library(copula) # extreme-value copulas
library(graphicalExtremes)
library(maps)
load("Temp.RData")
Temp <- as.data.frame(Temp)
```

 
# Temperature data in Belgium

Plotting a map of Belgium with the grid points for which we have temperature data:

```{r, fig.align="center"}
map("world", xlim = c(2,6.5), ylim = c(49,51.5)) 
# points(cbind(TempCoord[,"longitude"], TempCoord[,"latitude"]), col = "red")
text(cbind(TempCoord[,"longitude"], TempCoord[,"latitude"]))
```

Closest stations:

- Louvain-la-Neuve: around 30
- Leuven: around 20


# Some elementary univariate analysis 

The temperature data are annual maxima, so it makes sense to fit a GEV distribution to them. We do so for every grid point. The number of observations (rows) and number of grid points (columns):

```{r}
dim(Temp)
```

Fitting the GEV distribution to the annual temperature maxima in each grid point:

```{r}
# helper function to apply fit.gev to each column in a data.frame
fit.gev.vec <- function(data) {
  t(apply(data, 2, function(x) fit.gev(x)$estimate))
}
# fit GEV distribution to each 
TempGEVestim <- fit.gev.vec(Temp)
summary(TempGEVestim)
```

# Four cities in Belgium

```{r Temp.locations, fig.align="center"}
cities <- as.data.frame(rbind(
  c("Middelkerke", 1),
  c("Leuven", 20),
  c("LLN", 30),
  c("Virton", 54)
))
names(cities) <- c("name", "ID")
cities$ID <- as.integer(cities$ID)
TempPart <- Temp[,cities$ID]
names(TempPart) <- cities$name
map("world", xlim = c(2,6.5), ylim = c(49,51.5)) 
text(TempCoord[cities$ID,], col = "red", labels = cities$name)
```

```{r Tem.scatter, fig.align="center"}
plot(TempPart, main = "Maximum temperatures in Belgium, 1950—2018")
```

Fitting univariate GEV distributions to the annual maxima data in each city:

```{r Temp.GEV1}
Temp.GEV1 <- t(apply(TempPart, 2, function(x) fit.gev(x)$estimate))
Temp.GEV1
```

Checking the goodness-of-fit:

```{r fig.align="center"}
for (city in cities$name) {
  hist(TempPart[,city], freq = FALSE, 
     xlab = "Annual maximum temperature", 
     main = paste("Annual maximum temperatures at", city, "(1950—2014)"))
  curve(dgev(x, 
             loc = Temp.GEV1[city, "loc"], 
             scale = Temp.GEV1[city, "scale"], 
             shape = Temp.GEV1[city, "shape"]), 
        add = TRUE, col = "red")
  legend(x = "topleft", 
         col = c("black", "red"), 
         lty = "solid",
         legend = c("histogram", "GEV"))
}
```

# Pickands dependence functions

Nonparametric estimates of the bivariate Pickands dependence functions for each pair of cities:

```{r fig.align="center"}
for (i in 1:(nrow(cities)-1)) {
  for (j in (i+1):nrow(cities)) {
    curve(An.biv(TempPart[,c(i,j)], x, estimator = "CFG"),
          xlab = "w", ylab = "A(w)",
          main = paste("Pickands dependence function A,", 
                       cities$name[i], "vs", cities$name[j]),
          ylim = c(0.5,1),
          col = "blue", lwd = 2)
    lines(c(0, 0.5, 1), c(1,0.5,1), col = "gray")
    lines(c(0,1),c(1,1), col = "gray")
    Ahalf <- An.biv(TempPart[,c(i,j)], w=0.5)
    lines(c(0.5,0.5), c(1, Ahalf), lty = "dotted", lwd = 2)
    text(x = 0.5, y = (Ahalf+1)/2, 
         labels = paste("2{1-A(.5)} =", round(2*(1-Ahalf),3)),
         pos = 4
         )
  }
}
```


# Maximum pseudo-likelihood

Fitting a Gumbel copula (= logistic model) by maximum pseudo-likelihood, using ranks, and comparing the obtained Pickands dependence functions with the nonparametric estimates:

```{r fig.align="center"}
for (i in 1:(nrow(cities)-1)) {
  for (j in (i+1):nrow(cities)) {
    estimtmp <- fitCopula(copula = gumbelCopula(),
                          data = pobs(TempPart[, c(i,j)]))
    alpha <- summary(estimtmp)$coefficients[,"Estimate"]
    
    curve(An.biv(TempPart[,c(i,j)], x, estimator = "CFG"),
          xlab = "w", ylab = "A(w)",
          main = paste("Pickands dependence function A,", 
                       cities$name[i], "vs", cities$name[j]),
          ylim = c(0.5,1),
          col = "blue", lwd = 2)
    curve(A(gumbelCopula(param = alpha), x), add = TRUE, 
          col = "red", lwd = 2)
    lines(c(0, 0.5, 1), c(1,0.5,1), col = "gray")
    lines(c(0,1),c(1,1), col = "gray")
    legend(x = "top", lty = "solid", lwd = 2, col = c("blue", "red"),
           legend = c("nonparametric: CFG", 
                      paste("parametric: logistic, rho =", round(1/alpha,3))))
  }
}
rm(i,j,estimtmp,alpha)
```


# Parameter estimation

Fitting a parametric bivariate GEV (margins + dependence) to a pair of cities:

```{r}
library(evd)
(res.fbvevd.log <- fbvevd(x = Temp[,c(30,54)], model = "log"))
```


```{r}
(res.fbvevd.HR <- fbvevd(x = Temp[,c(30,54)], model = "hr"))
```