---
title: "Exercice session 1: univariate extremes"
author: "Anna Kiriliouk and Johan Segers"
date: "UCLouvain, October 2024"
output: html_document
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{mathtools}
  - \newcommand{\rbr}[1]{\left(#1\right)}
  - \newcommand{\cbr}[1]{\left\{#1\right\}}
  - \newcommand{\pr}{\operatorname{\mathsf{P}}}
  - \newcommand{\expec}{\operatorname{\mathsf{E}}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1. 

For this first exercise, you'll need to install the `ismev` package first using `install.packages("ismev")`. Next, load the package and the data. The `data` vector contains daily maximal speeds of wind gusts, measured in kilometers per hour in Eindhoven (the Netherlands) during extended winter (October–March), from October 2001 up to and including March 2022. 

```{r message = FALSE}
library(ismev)
load("Ex1uni.RData")
```

a. Use the following command to select monthly maxima. Then also select the yearly maxima (one for each extended winter season spanning from October to March).

```{r}
monthly <- as.vector(tapply(data, paste(times$year,times$mon), max))
yearly <- apply(matrix(monthly,ncol = 21, nrow = 6), 2, max)
```

b. Plot the two datasets. Do they look stationary?

```{r}
plot(monthly, ylim = c(35,130))
plot(yearly, ylim = c(35,130))
```

c. Fit a GEV distribution to both the monthly and the yearly maxima and check the goodness-of-fit plots. You'll need to use the functions `gev.fit`, and `gev.diag`. Which one of the two would you use?

```{r}
gmonthly <- gev.fit(monthly, show = FALSE)
gyearly <- gev.fit(yearly, show = FALSE)
gev.diag(gmonthly)
gev.diag(gyearly)
# I would go with the monthly maxima: the fit seems satisfactory and confidence intervals would be lower. 
```

d. For the chosen data, give the GEV parameter estimates with 95 \% confidence intervals. Is the data bounded-tailed, light-tailed or heavy-tailed? 
```{r}
round(cbind(gmonthly$mle - 1.96*gmonthly$se, gmonthly$mle, gmonthly$mle + 1.96*gmonthly$se),2)
round(cbind(gyearly$mle - 1.96*gyearly$se, gyearly$mle, gyearly$mle + 1.96*gyearly$se),2)
# Light-tailed, in the Gumbel domain of attraction
```

e. Using the function `gum.fit`, estimate the parameters of a GEV distribution with $\xi = 0$. Decide whether such a model suffices based on a likelihood ratio test. *Reminder: if $L(\xi,\sigma,\mu)$ denotes the log-likelihood function, then $2 \left\{L(\xi,\sigma,\mu) - L(0,\sigma,\mu)\right\} \rightarrow \chi^2_1$.*

```{r}
gummonthly <- gum.fit(monthly, show = FALSE)
gumyearly <- gum.fit(yearly, show = FALSE)
gum.diag(gummonthly)
gum.diag(gumyearly)
(Dmonthly <- 2*(gummonthly$nllh - gmonthly$nllh))
(Dyearly <- 2*(gumyearly$nllh - gyearly$nllh))
qchisq(0.95, df = 1)
```

f. The return-level plot returned by `gum.diag` is not very clear. Using the formula seen this morning (slide nr),
calculate the 10-year and the 1000-year return levels. Confidence intervals can be obtained using the delta method: if $\xi = 0$, $\nabla x_p^T = (1, - \log(-\log(1-p)))$, and the covariance matrix of $(\hat{\mu},\hat{\xi})$ can be obtained from the output of `gum.fit`. Is there a big difference between the 10-year and the 1000-year return levels? Is this surprising?

```{r}
retLevGum <- function(pars, cov, p){
  rl <- pars[1] - pars[2]*log(-log(1-p))
  grd <- c(1, -log(-log(1-p)))
  se <- sqrt(t(grd) %*% cov %*% grd)
  return(c(rl - 1.96*se, rl, rl + 1.96*se))
}
retLevGum(gummonthly$mle, gummonthly$cov, (1/10)/6) #convert to yearly
retLevGum(gummonthly$mle, gummonthly$cov, (1/1000)/6) #convert to yearly
retLevGum(gumyearly$mle, gumyearly$cov, 1/10)
retLevGum(gumyearly$mle, gumyearly$cov, 1/1000)
```


