---
title: "Multivariate maxima"
author: "Anna Kiriliouk and Johan Segers"
date: "UCLouvain and KU Leuven, October 2024"
output: html_document
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{bm}
  - \newcommand{\rbr}[1]{\left(#1\right)}
  - \newcommand{\cbr}[1]{\left\{#1\right\}}
  - \newcommand{\pr}{\operatorname{\mathsf{P}}}
  - \newcommand{\expec}{\operatorname{\mathsf{E}}}
  - \newcommand{\sig}{\sigma}
  - \newcommand{\evi}{\xi}
  - \newcommand{\bevi}{\bm{\evi}}
  - \newcommand{\stdf}{L}%{\ell}
  - \newcommand{\Pick}{A}
  - \newcommand{\exCoef}{\theta}
  - \newcommand{\expms}{\Lambda}
  - \newcommand{\angms}{\varPhi}
  - \newcommand{\angmspdf}{\phi}
  - \newcommand{\hstdf}{\widehat{\stdf}}
  - \newcommand{\hangms}{\widehat{\angms}}
  - \newcommand{\hbP}{\widehat{\bP}}
  - \newcommand{\hP}{\widehat{P}}
  - \newcommand{\hR}{\widehat{R}}
  - \newcommand{\hF}{\widehat{F}}
  - \newcommand{\hFj}{\widehat{F}_{j}}
  - \newcommand{\rank}{\operatorname{Rnk}}
  - \newcommand{\hU}{\widehat{U}}
  - \newcommand{\hpr}{\widehat{\pr}}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r prelim}
library(mev)
library(copula) # extreme-value copulas
library(graphicalExtremes)
library(maps)
load("Wind.RData")
load("Temp.RData")
Temp <- as.data.frame(Temp)
```

 
----------


# Things to show and explain

## Objects to introduce

Dependence functions: $\ell$, $A$

Dependence coefficients: $\theta$, $\chi$

Measures: angular measure $H$, exponent measure (?)

Much more--what to choose \ldots?



## Nonparametric inference

- Bivariate Pickands dependence function
- Calculate pairwise $\chi$
- test of max-stability (?)
- Angular measure (also for multivariate maxima?)

## Parametric inference

- Fit parametric models
- Simulate


# Maps

## Temperature data in Belgium

Plotting a map of Belgium with the grid points for which we have temperature data:

```{r}
map("world", xlim = c(2,6.5), ylim = c(49,51.5)) 
# points(cbind(TempCoord[,"longitude"], TempCoord[,"latitude"]), col = "red")
text(cbind(TempCoord[,"longitude"], TempCoord[,"latitude"]))
```

Closest stations:
- Louvain-la-Neuve: around 30
- Leuven: around 20

## Wind data in the Netherlands

Plotting a map of the Netherlands with the weather stations for which we have wind data:

```{r}
map("world", xlim = c(3.2,7.5), ylim = c(50.75,54)) 
points(WindCoord, col = "red")
```

# Some elementary univariate analysis 

## Temperature data

The temperature data are annual maxima, so it makes sense to fit a GEV distribution to it. We do so for every grid point. The number of observations (rows) and number of grid points (columns):

```{r}
dim(Temp)
```

Fitting the GEV distribution to the annual temperature maxima in each grid point:

```{r}
# helper function to apply fit.gev to each column in a data.frame
fit.gev.vec <- function(data) {
  t(apply(data, 2, function(x) fit.gev(x)$estimate))
}
# fit GEV distribution to each 
TempGEVestim <- fit.gev.vec(Temp)
summary(TempGEVestim)
```

## Wind data

Sample size and number of variables:

```{r}
dim(Wind)
```

Fitting a univariate GEV distribution to the first station:

```{r}
fit.gev(Wind[,1], show=TRUE)
```

# Nonparametric multivariate analysis

## Plotting bivariate data

```{r Temp.locations}
cities <- as.data.frame(rbind(
  c("Middelkerke", 1),
  c("Leuven", 20),
  c("LLN", 30),
  c("Virton", 54)
))
names(cities) <- c("name", "ID")
cities$ID <- as.integer(cities$ID)
TempPart <- Temp[,cities$ID]
names(TempPart) <- cities$name
map("world", xlim = c(2,6.5), ylim = c(49,51.5)) 
text(TempCoord[cities$ID,], col = "red", labels = cities$name)
```

```{r Tem.scatter}
plot(TempPart, main = "Maximum temperatures in Belgium, 1950—2018")
```

```{r Temp.GEV1}
Temp.GEV1 <- t(apply(TempPart, 2, function(x) fit.gev(x)$estimate))
Temp.GEV1
```

```{r}
for (city in cities$name) {
  hist(TempPart[,city], freq = FALSE, 
     xlab = "Annual maximum temperature", 
     main = paste("Annual maximum temperatures at", city, "(1950—2014)"))
  curve(dgev(x, 
             loc = Temp.GEV1[city, "loc"], 
             scale = Temp.GEV1[city, "scale"], 
             shape = Temp.GEV1[city, "shape"]), 
        add = TRUE, col = "red")
  legend(x = "topleft", 
         col = c("black", "red"), 
         lty = "solid",
         legend = c("histogram", "GEV"))
}
```

## Pickands dependence functions

```{r}
for (i in 1:(nrow(cities)-1)) {
  for (j in (i+1):nrow(cities)) {
    curve(An.biv(TempPart[,c(i,j)], x, estimator = "CFG"),
          xlab = "w", ylab = "A(w)",
          main = paste("Pickands dependence function A,", 
                       cities$name[i], "vs", cities$name[j]),
          ylim = c(0.5,1),
          col = "blue", lwd = 2)
    lines(c(0, 0.5, 1), c(1,0.5,1), col = "gray")
    lines(c(0,1),c(1,1), col = "gray")
    Ahalf <- An.biv(TempPart[,c(i,j)], w=0.5)
    lines(c(0.5,0.5), c(1, Ahalf), lty = "dotted", lwd = 2)
    text(x = 0.5, y = (Ahalf+1)/2, 
         labels = paste("2{1-A(.5)} =", round(2*(1-Ahalf),3)),
         pos = 4
         )
  }
}
```

## Testing max-stability

Based on Kojadinovic and Yan (2010):

```{r}
ncitypairs <- nrow(cities) * (nrow(cities)-1) / 2
testoutA <- data.frame(matrix(0, nrow = ncitypairs, ncol = 4))
names(testoutA) <- c("city1", "city2", "statistic", "p.value")
testoutA$city1 <- as.character(testout$city1)
testoutA$city2 <- as.character(testout$city2)

counter <- 0
for (i in 1:(nrow(cities)-1)) {
  for (j in (i+1):nrow(cities)) {
    counter <- counter +1
    testoutA$city1[counter] <- cities$name[i]
    testoutA$city2[counter] <- cities$name[j]
    testoutij <- evTestA(x = as.matrix(TempPart[,c(i,j)]), 
                         derivatives = "Cn")
    testoutA$statistic[counter] <- testoutij$statistic
    testoutA$p.value[counter] <- testoutij$p.value
  }
}
rm(counter, testoutij)
testoutA
```

Other tests implemented in `copula` package: `evTestK` and `evTestC`.

# Parameter estimation

## Composite likelihood

Function `ExtremalDep::fExtDep()`. Not sure what it does, how margins are taken into account...

```{r}
library(ExtremalDep)
fExtDep(method="Composite", data = as.matrix(Temp[, c(1,2,3)]), model = "HR", par.start = c(1,1,1))
```

```{r}
Temp.pobsFr <- - 1 / log(pobs(Temp))
fExtDep(method="Composite", data = as.matrix(Temp.pobsFr[, c(1,2,3)]), model = "HR", par.start = c(1,1,1))
```

Function `SpatialExtremes::fitmaxstab()`: R session crashes...

```{r}
#fitmaxstab(data = as.matrix(Temp[,c(1,20,30,54)]),
#           coord = as.matrix(TempCoord[c(1,20,30,54),]),
#           cov.mod = "gauss",
#           fit.marge = TRUE)
```

```{r}
library(SpatialExtremes)
library(copula)
load("Temp.RData")
location.IDs <- c(1,20,54)
fitmaxstab(data = -1/log(pobs(as.matrix(Temp[,location.IDs]))),
           coord = as.matrix(TempCoord[location.IDs,]),
           cov.mod = "gauss",
           fit.marge = FALSE)$fitted.values
```


```{r}
location.IDs <- c(1,54)
require(SpatialExtremes)
fitmaxstab(data = as.matrix(Temp[,location.IDs]),
           coord = as.matrix(TempCoord[location.IDs,]),
           cov.mod = "gauss",
           fit.marge = TRUE)$fitted.values
```

Compare multivariate fit with three univariate fits:

```{r}
library(mev)
t(apply(Temp[,location.IDs], 2, function(x) fit.gev(x)$estimate))
```

## parametric

```{r}
library(evd)
load("Temp.RData")
res.fbvevd <- fbvevd(x = Temp[,c(1,20)], model = "log")
res.fbvevd
```

```{r}
res.fbvevd.HR <- fbvevd(x = Temp[,c(1,54)], model = "hr")
res.fbvevd.HR
```

```{r}
res.fbvevd.log <- fbvevd(x = Temp[,c(1,54)], model = "log")
res.fbvevd.log
```

```{r}
res.fbvevd.alog <- fbvevd(x = Temp[,c(1,54)], start = as.list(c(asy1=1,asy2=1,dep=0.54711)), model = "alog")
res.fbvevd.alog
```

```{r}
library(ismev)
gev.fit(Temp[,1])
gev.fit(Temp[,20])
```

## Maximum pseudo-likelihood

Fitting a Gumbel copula (= logistic model) by maximum pseudo-likelihood, using ranks, and comparing the obtained Pickands dependence functions:

```{r}
for (i in 1:(nrow(cities)-1)) {
  for (j in (i+1):nrow(cities)) {
    estimtmp <- fitCopula(copula = gumbelCopula(),
                          data = pobs(TempPart[, c(i,j)]))
    alpha <- summary(estimtmp)$coefficients[,"Estimate"]
    
    curve(An.biv(TempPart[,c(i,j)], x, estimator = "CFG"),
          xlab = "w", ylab = "A(w)",
          main = paste("Pickands dependence function A,", 
                       cities$name[i], "vs", cities$name[j]),
          ylim = c(0.5,1),
          col = "blue", lwd = 2)
    curve(A(gumbelCopula(param = alpha), x), add = TRUE, 
          col = "red", lwd = 2)
    lines(c(0, 0.5, 1), c(1,0.5,1), col = "gray")
    lines(c(0,1),c(1,1), col = "gray")
    legend(x = "top", lty = "solid", lwd = 2, col = c("blue", "red"),
           legend = c("nonparametric: CFG", 
                      paste("parametric: logistic, rho =", round(1/alpha,3))))
  }
}
rm(i,j,estimtmp,alpha)
```

## Testing a parametric model for the GEV dependence structure

Testing the Gumbel copula (= logistic model) based on pseudo-observations, estimating copula parameter with maximum pseudo-likelihood:

```{r goodness.of.fit, eval=FALSE}
# Code takes some time to run; set eval=FALSE to execute it
gofCopula(copula = gumbelCopula(),
          x = as.matrix(TempPart[,c("Leuven", "Virton")]),
          ties = TRUE,
          verbose = FALSE)
```


## Dependence coefficients

Using `mev` package, for one pair of variables:

```{r}
chi <- taildep(data = Wind[,1:2], depmeas = "chi", method = list(chi = "emp"))
chi
```

Using `emp_chi()` from `graphicalExtremes` package, giving a $d \times d$ matrix:

```{r}
chi_hat <- emp_chi(data = Wind, p = .8)
```

The extremal correlation $\chi \in [0, 1]$, the extremal coefficient $\theta \in [1,2]$, the stdf $\ell : [0, \infty)^2 \to [0, \infty)$ and the Pdf $A : [0,1] \to [1/2,1]$ are related by
\[
  \chi = 2 - \theta \quad \text{and} \quad 
  \theta = \ell(1,1) = 2 A(1/2).
\]
For a bivariate max-stable random vector $(Z_1,Z_2)$, with unit-Fréchet margins, we have
\[
  \pr \rbr{ 1/Z_1 > x, 1/Z_2 > y }
  = \exp \cbr{ -(x+y) \, A \rbr{\frac{y}{x+y}}}, \qquad x,y > 0.
\]
It follows that
\[
  \pr \cbr{ 1/\max(Z_1,Z_2) > x } = \exp \cbr{-2x A(1/2)} = \exp(-\theta x), 
  \qquad x > 0
\]
and thus 
\[ 
  \expec \cbr{\min(1/Z_1,1/Z_2)} = \expec \cbr{1/\max(Z_1,Z_2)} = 1 / \theta. 
\]
Given $d$-variate block maxima data $M_t = (M_{t1},\ldots,M_{td})$ for $t = 1,\ldots,n$, we can thus estimate these quantities for variables $i,j \in \cbr{1,\ldots,d}$ by
\begin{align*}
  \hat{\chi}_{ij} &= 2 - \hat{\theta}_{ij} \\
  \frac{1}{\hat{\theta}_{ij}} 
  &= \frac{1}{n} \sum_{t=1}^n \min \cbr{ \hat{E}_{ti}, \hat{E}_{tj} }, \\
  \hat{E}_{ti} &= 1/\hat{Z}_{ti} = - \log \hat{F}_i(M_{ti}),
\end{align*}
where $\hat{F}_i$ is an estimate (parametric or non-parametric) of the marginal cdf in margin $i$.


```{r}
emp_chi_max <- function(data) {
  d <- ncol(data)
  if (d < 2) return(1) # data are univariate
  n <- nrow(data)
  rankdata <- apply(data, 2, rank)
  Z <- log(n+1) - log(rankdata)
  chi_hat <- diag(d)
  for (i in 1:(d-1)) {
    for (j in (i+1):d) {
      chi_hat[i,j] <- chi_hat[j,i] <- 2-1/mean(pmin(Z[,i], Z[,j]))
    }
  }
  return(chi_hat)
}
chi_hat <- emp_chi_max(data = Wind)
chi_hat[1:5,1:5]
```



Transform observations component-wise to ranks, transform to unit-Fréchet margins, and estimate 

```{r}
WindRanks <- apply(Wind, MARGIN = 2, FUN = rank)
dim(WindRanks)
```

